```{r libraries, echo = FALSE, include = FALSE}
require("ggplot2")
require("ggokabeito")
require("tidyr")
require("dplyr")
require("stringr")
require("knitr")
require("pander")
require("patchwork")
require("latex2exp")
```

```{r setup, echo = FALSE}
options(dplyr.summarise.inform = FALSE)
knitr::opts_knit$set(kable.force.latex = TRUE)

meanse <- function(x, ...) {
    mean1 <- signif(round(mean(x, na.rm = T), 2), 5) # calculate mean and round
    se1 <- signif(round(sd(x, na.rm = T) / sqrt(sum(!is.na(x))), 2), 2) # std error - round adding zeros
    out <- paste(mean1, "$\\pm$", se1) # paste together mean plus/minus and standard error
    if (str_detect(out, "NA")) {
        out <- "NA"
    } # if missing do not add plusminus
    if (se1 == 0) {
        out <- paste(mean1)
    }
    return(out)
}
```

```{r import-data, echo = FALSE}
filenames <- list.files(
    path = "./data/",
    pattern = "inference_testing_2023-07-*",
    full.names = TRUE
)
data_list <- lapply(filenames, readRDS)
col_id_BICLS <- c(11, 16, 23, 30, 37)
result_data_frame <- dplyr::bind_rows(data_list)

# Compute the preferred model
result_data_frame <- cbind(result_data_frame, preferred_model = sapply(seq_len(nrow(result_data_frame)), function(n) sub("_BICL", "", names(which.max(result_data_frame[n, col_id_BICLS])))))
result_data_frame$preferred_model <- factor(result_data_frame$preferred_model, levels = c(
    "sep", "iid", "pi",
    "rho", "pirho"
))
```

# Efficiency of the inference

\paragraph{Simulation settings} For this simulation the data is simulated with 
$M = 2, n_{1}^{m} = 120, n_{2}^{m} = 120, Q_1 = Q_2 = 4$, $\bm{\alpha}, \bm{\pi}$
and $\bm{\rho}$ are set as follows:
\begin{align*}
    &&\bm{\alpha} = .25 + 
                    \begin{pmatrix}
                        3 \eps[\alpha] & 2 \eps[\alpha] & \eps[\alpha] & - \eps[\alpha]\\
                        2 \eps[\alpha] & 2 \eps[\alpha] & - \eps[\alpha] & \eps[\alpha]\\
                        \eps[\alpha] & - \eps[\alpha] & \eps[\alpha] & 2 \eps[\alpha]\\
                        - \eps[\alpha] & \eps[\alpha] & 2 \eps[\alpha] & 0
                    \end{pmatrix}, \\ \bm{\pi}^1 = \sigma_1 
                    \begin{pmatrix}
                        0.2 & 0.4 & 0.4 & 0
                    \end{pmatrix},
                    && \bm{\pi}^2 = 
                    \begin{pmatrix}
                        0.25 & 0.25 & 0.25 & 0.25
                    \end{pmatrix}, \\
                    \bm{\rho}^1 = 
                    \begin{pmatrix}
                        0.25 & 0.25 & 0.25 & 0.25
                    \end{pmatrix}, &&
                    \bm{\rho}^2 = \sigma_2
                    \begin{pmatrix}
                        0 & 0.33 & 0.33 & 0.33
                    \end{pmatrix}, &&
\end{align*}
with $\eps[\alpha]$ taking nine equally spaced values ranging from 0 to 0.24.
For each value of $\eps[\alpha]$, 108 datasets ($X_1, X_2$) are simulated, 
resulting in $9 \times 108 = 972$ datasets. More precisely, for each dataset, 
we pick uniformly at random two permutations of $\{ 1, \dots , 4 \}$ 
($\sigma_1, \sigma_2$) with the constraint that $\sigma_1(4) \neq \sigma_2(1)$. 
This ensures that each of the two networks have a non-empty block that is empty 
in the other one. Then the networks are simulated with 
$\mathcal{B}$ern-$BiSBM_{120}(4, \bm{\alpha}, \bm{\pi}^m, \bm{\rho}^m)$
with the previous parameters. Each network has 2 blocks in common and their 
connectivity structures encompass a mix of core-periphery, assortative 
community and disassortative community structures, depending on which 3 of the 4
blocks are selected for each network. $\eps[\alpha]$ represents the strength of
these structures, the larger, the easier it is to tell apart one block from 
another.
The true model of all the simulation is a $\pi\rho\text{-}colBiSBM$.

\paragraph{Inference} We want to measure the quality of the
inference procedure, for this we use the inference described in the section
\ref{sec:variational-estimation-of-the-parameters}.

\paragraph{Quality indicators} To assess the quality of the inference, we will
use the following indicators:
\begin{itemize}
    \item First, for each dataset, we put in competition $\pi\text{-}colBiSBM$ with 
    $sep\text{-}BiSBM$, $iid\text{-}colBiSBM$, $\rho\text{-}colBiSBM$, 
    $\pi\rho\text{-}colBiSBM$
    respectively. To do so, for each dataset, we compute the 
    BIC-L of each model $\pi\text{-}colBiSBM$ is preferred to $sep\text{-}BiSBM$
    (resp. $iid\text{-}colBiSBM$, $\rho\text{-}colBiSBM$, 
    $\pi\rho\text{-}colBiSBM$) if 
    its BIC-L is greater.
    \item When considering $\pi\text{-}colBiSBM$, $\rho\text{-}colBiSBM$, 
    $\pi\rho\text{-}colBiSBM$ we compare $\widehat{Q_1}$, $\widehat{Q_2}$ to 
    their true values. ($Q_1 = 4$ and $Q_2 = 4$)
    \item Finally, we assess the quality of the node grouping by computing the
    Adjusted Rand Index \parencite[][, ARI = 0 for a random grouping, ARI = 1 for a perfect recovery]{hubertComparingPartitions1985}. For each network, for the
    $\pi\text{-}colBiSBM$, $\rho\text{-}colBiSBM$, 
    $\pi\rho\text{-}colBiSBM$ we compare the inferred block memberships to the
    real ones by computing the mean of the ARI per axis over the two networks
    \begin{equation*}
        \overline{\text{ARI}}_d = \frac{1}{2} \text{ARI}\big( \text{ARI}(\widehat{\bm{Z}^1_d},\bm{Z}^1_d) + \text{ARI}(\widehat{\bm{Z}^2_d},\bm{Z}^2_d) \big) 
    \end{equation*}
    where $d$ is the dimension or axis (i.e., rows, $d=1$, or columns, $d=2$) of
    the block memberships.
    And we compute the ARI of the whole set of nodes to account for block
    pairing between networks
    \begin{equation*}
        \text{ARI}_d = \text{ARI}\big((\widehat{\bm{Z}^1_d},\widehat{\bm{Z}^2_d}),(\bm{Z}^1_d,\bm{Z}^2_d) \big) 
    \end{equation*}
\end{itemize} 

All these quality indicators are averaged over the 108 datasets. The results are
provided in the tables \ref{tab:per_model_sep} to \ref{tab:per_model_pirho}. Each line corresponds to the 
108 datasets for a given value of value of $\eps[\alpha]$. 

```{r inference_table, echo = FALSE}
averaged_data <- result_data_frame %>%
    group_by(epsilon_alpha) %>%
    summarise(across(-preferred_model, list("avrg" = meanse))) %>%
    select(-c(2:10))
averaged_data <- averaged_data %>%
    select(which(!grepl("*_BICL_*", colnames(averaged_data)), 
    arr.ind = TRUE))
```

```{r function_per_model, echo = FALSE}
dataframe_per_model <- function(model) {
    averaged_data %>%
        select(epsilon_alpha, starts_with(paste0(model, "_")))
}
```

\tiny
```{r per_model_table, echo = FALSE, results='asis', message=FALSE, warning = FALSE}
for (model in c("sep", "iid", "pi", "rho", "pirho")) {
    kable_colnames <- c(
        "$\\eps[\\alpha]$", #"BIC-L", 
        "$\\overline{\\text{ARI}}_{1}$",
        "$\\overline{\\text{ARI}}_{2}$", "$\\text{ARI}_{1}$", "$\\text{ARI}_{2}$"
    )
    model_name <- model
    if (model != "sep") {
        kable_colnames <- c(
            kable_colnames, "Recovered $Q_1$",
            "Recovered $Q_2$"
        )
    }
    if (model == "pirho") {
        model_name <- "$\\pi\\rho$"
    } else {
        if (model != "iid" && model != "sep") {
            model_name <- paste0("$\\", model, "$")
        } else {
            model_name <- paste0("$", model, "$")
        }
    }
    print(kable(dataframe_per_model(model),
        escape = FALSE,
        booktabs = TRUE,
        digits = 2,
        position = "!h",
        caption = paste0(
            "\\label{tab:per_model_", model,
            "}Quality metrics for ",
            ifelse(model != "sep", paste0(model_name, "$\\text{-}colBiSBM$"),"$sep\\text{-}BiSBM$")
        ),
        col.names = kable_colnames
    ))
}
```
\normalsize

```{r proportion-preferred_model, echo = FALSE}
proportion_preferred_data <- result_data_frame %>%
    group_by(epsilon_alpha, preferred_model) %>%
    summarise(n = n()) %>%
    mutate(prop_model = n / sum(n)) %>%
    ungroup() %>%
    select(-n)

proportion_preferred_table <- proportion_preferred_data %>%
    pivot_wider(
        names_from = preferred_model,
        values_from = prop_model, values_fill = 0
    )

kable(proportion_preferred_table,
    escape = FALSE,
    booktabs = TRUE,
    digits = 2,
    position = "!h",
    caption = "\\label{tab:proportion-preferred-table}Proportions of models selected per \\eps[\\alpha] (data for Figure \\ref{fig:inference-proportion-preferred})",
    col.names = c(
        "\\eps[\\alpha]",
        "$sep\\text{-}BiSBM$",
        "$iid\\text{-}colBiSBM$",
        "$\\pi\\text{-}colBiSBM$",
        "$\\rho\\text{-}colBiSBM$",
        "$\\pi\\rho\\text{-}colBiSBM$"
    ),
    align = "rccccc",
    format = "latex"
)
```
```{r proportion_preferred_figure, echo = FALSE}
#| fig.cap="\\label{fig:inference-proportion-preferred}Plot of the proportions of different preferred models in function of \\eps[\\alpha]",
#| fig.asp = 0.5,
#| fig.pos = "H",
#| fig.width = 7,
#| fig.height = 4,
#| dpi=300

proportion_preferred_data %>% ggplot() +
    aes(
        x = epsilon_alpha, y = prop_model, color = preferred_model,
        fill = preferred_model
    ) +
    guides(
        fill = guide_legend(title = "Preferred Model"),
        color = guide_legend(title = "Preferred Model")
    ) +
    scale_x_continuous(breaks = seq(from = 0.0, to = 0.24, by = 0.03)) +
    scale_color_okabe_ito() +
    scale_fill_okabe_ito() +
    xlab(TeX("$\\epsilon_{\\alpha}$")) +
        ylab("Model proportions") +
        geom_col(position = "stack")

```

\paragraph{Results} For the model comparison, when $\eps[\alpha]$ is small 
($\eps[\alpha]\in[0, .04]$), the simulation model is close to the 
Erd\H{o}s-Reńyi network and it is very hard to find any structure beyond the one
of a single block on each dimension.

On the figure \ref{fig:inference-proportion-preferred} and table 
\ref{tab:proportion-preferred-table} we can see that from
$\eps[\alpha] = 0.12$ around $70\%$ of the time the $\pi\rho\text{-}colBiSBM$
model (i.e., the correct one) is selected. 

An interesting result we can read in the tables is that our models outperform
the $sep\text{-}BiSBM$ when considering the ARI on the whole set of nodes 
($\text{ARI}_d$). This means that our models are able to recover the block 
pairing \emph{between the networks} in addition to recovering the blocks and
their parameters.