```{r libraries, echo = FALSE, include = FALSE}
require("ggplot2")
require("ggokabeito")
require("knitr")
require("kableExtra")
require("stringr")
require("tidyr")
require("dplyr")
require("patchwork")
require("latex2exp")
```

```{r setup, echo = FALSE, include= FALSE}
options(knitr.table.format = "latex")

meanse <- function(x, ...) {
    mean1 <- signif(round(mean(x, na.rm = T), 2), 5) # calculate mean and round
    se1 <- signif(round(sd(x, na.rm = T) / sqrt(sum(!is.na(x))), 2), 2) # std error - round adding zeros
    out <- paste(mean1, "$\\pm$", se1) # paste together mean plus/minus and standard error
    if (str_detect(out, "NA")) {
        out <- "NA"
    } # if missing do not add plusminus
    if (se1 == 0) {
        out <- paste(mean1)
    }
    return(out)
}
```


```{r import-data, echo = FALSE}
filenames <- list.files(
    path = "./data/",
    pattern = "model_selection_check_batch_report_3_rep_*",
    full.names = TRUE
)
data_list <- lapply(filenames, readRDS)
result_data_frame <- dplyr::bind_rows(data_list)

result_data_frame$preferred_model <- factor(result_data_frame$preferred_model, levels = c(
    "sep","iid","pi",
    "rho","pirho"
))

# Adding a column accounting for true model iid, pi, rho or pirho
# result_data_frame <- result_data_frame %>% mutate(true_model = if (all( c(epsilon_pi >0, epsilon_rho > 0) == c(TRUE, TRUE))) print("pirho") else if (all( c(epsilon_pi >0, epsilon_rho > 0) == c(TRUE, FALSE))) print("pi") else if (all( c(epsilon_pi >0, epsilon_rho > 0) == c(F, T))) print("rho") else print("iid"))

```

# Capacity to distinguish $\pi\rho\text{-}colBiSBM$ from $iid\text{-}colBiSBM$ and other variants

The idea of this model selection simulations is to assess how the model select 
the correct *colBiSBM* model among the possible ones: 
\textit{iid, pi, rho, pirho}. This difference being based on the row and col
block proportions.

For this task we choose the same simulation context
as \cite{chabert-liddellLearningCommonStructures2023}.

Namely $n_{1}^{m} = 90, n_{2}^{m} = 90, Q_1 = Q_2 = 3$, $\bm{\alpha}, \bm{\pi}$
and $\bm{\rho}$ are set as follows:

\begin{align*}
    \bm{\alpha} =.25 +  \begin{pmatrix}
                        3 \eps[\alpha] & 2 \eps[\alpha] & \eps[\alpha] \\
                        2 \eps[\alpha] & 2 \eps[\alpha] & - \eps[\alpha] \\
                        \eps[\alpha] & - \eps[\alpha] & \eps[\alpha]
                    \end{pmatrix}, & & \bm{\pi}^1 = \begin{pmatrix}
                        \frac{1}{3}, & \frac{1}{3}, & \frac{1}{3}
                    \end{pmatrix}, & & \bm{\pi}^2 = \sigma\begin{pmatrix}
                        \frac{1}{3} - \eps[\pi], & \frac{1}{3}, & \frac{1}{3} + \eps[\pi]
                    \end{pmatrix},\\
    & & \bm{\rho}^1 = \begin{pmatrix}
                        \frac{1}{3}, & \frac{1}{3}, & \frac{1}{3}
                    \end{pmatrix}, & & \bm{\rho}^2 = \sigma\begin{pmatrix}
                        \frac{1}{3} - \eps[\rho], & \frac{1}{3}, & \frac{1}{3} + \eps[\rho]
                    \end{pmatrix},
\end{align*}
with $\eps[\alpha] = 0.16$, $\eps[\pi]$ and $\eps[\rho]$ taking 9 values equally
spaced in $\left[ 0, .28\right]$. We simulate 324 different collections for each
value of $\eps[\pi]$ and $\eps[\rho]$.

$\pi\rho\text{-}colBiSBM$, $\pi\text{-}colBiSBM$, $\rho\text{-}colBiSBM$,
$iid\text{-}colBiSBM$ and $sep\text{-}BiSBM$ are put in competition and the
model with the greater BIC-L is selected as the \emph{preferred model}. 

```{r compute-table, echo = FALSE, include = FALSE}
model_comparison_eps_pi_rho <- result_data_frame %>%
    group_by(epsilon_pi, epsilon_rho, preferred_model) %>%
    summarise(n = n()) %>%
    mutate(prop_model = n / sum(n))

model_comparison_eps_pi <- result_data_frame %>%
    group_by(epsilon_pi, preferred_model) %>%
    summarise(n = n(), rec_Q1 = mean(iid_Q1 + pi_Q1 + rho_Q1 + pirho_Q1)/4) %>%
    mutate(prop_model = n / sum(n))

model_comparison_eps_rho <- result_data_frame %>%
    group_by(epsilon_rho, preferred_model) %>%
    summarise(n = n(), rec_Q2 = mean(iid_Q2 + pi_Q2 + rho_Q2 + pirho_Q2)/4) %>%
    mutate(prop_model = n / sum(n))
```

```{r epsilon_plot, echo = FALSE, include = FALSE}
#| fig.asp = 0.5,
#| fig.pos = "H",
#| fig.width = 7,
#| fig.height = 4,
#| dpi=300

plot_pi <- model_comparison_eps_pi %>% ggplot() +
    aes(
        x = epsilon_pi, y = prop_model, color = preferred_model,
        fill = preferred_model
    ) +
    guides(fill = guide_legend(title = "Preferred Model"),
    color = guide_legend(title = "Preferred Model")) +
    xlab(TeX("$\\epsilon_{\\pi}$")) +
    ylab("Model proportions") +
    scale_color_okabe_ito() +
    scale_fill_okabe_ito() +
    geom_col(position = "stack")
plot_rho <- model_comparison_eps_rho %>% ggplot() +
    aes(
        x = epsilon_rho, y = prop_model, color = preferred_model,
        fill = preferred_model
    ) +
    guides(fill = guide_legend(title = "Preferred Model"),
    color = guide_legend(title = "Preferred Model")) +
    xlab(TeX("$\\epsilon_{\\rho}$")) +
    ylab("") +
    scale_color_okabe_ito() +
    scale_fill_okabe_ito() +
    geom_col(position = "stack")

ggsave("./img/plot_model_function_eps.png", plot_pi + plot_rho +
    plot_layout(guides = "collect"))
```


When $\eps[\pi] = 0$, $\bm{\pi}^1 = \bm{\pi}^2$, $\eps[\rho] = 0$ and
$\bm{\rho}^1 = \bm{\rho}^2$, the generated collection is an 
$iid\text{-}colBiSBM$. When $\eps[\pi] > 0$ or $\bm{\pi}^1 \neq \bm{\pi}^2$,
the model is a $\pi\text{-}colBiSBM$.
When $\eps[\rho] > 0$ or $\bm{\rho}^1 \neq \bm{\rho}^2$,
the model is a $\rho\text{-}colBiSBM$. 
Finally, when $\eps[\pi] > 0$ or $\bm{\pi}^1 \neq \bm{\pi}^2$ and
$\eps[\rho] > 0$ or $\bm{\rho}^1 \neq \bm{\rho}^2$,
the model is a $\pi\rho\text{-}colBiSBM$.

```{r tables, echo = FALSE, results='asis'}
kable(
    (model_comparison_eps_pi %>%
        select(-one_of("n")) %>%
        pivot_wider(
            names_from = preferred_model,
            values_from = prop_model,
            values_fill = 0
        ) %>% group_by(epsilon_pi) %>%
        summarise(
            rec_Q1 = meanse(rec_Q1),
            iid = sum(iid),
            pi = sum(pi),
            rho = sum(rho),
            pirho = sum(pirho)
        ))[, c(1, 3:6, 2)],
    digits = 2,
    col.names = c(
        "$\\eps[\\pi]$",
        "$iid\\text{-}colBiSBM$    ",
        "$\\pi\\text{-}colBiSBM$",
        "$\\rho\\text{-}colBiSBM$",
        "$\\pi\\rho\\text{-}colBiSBM$",
        "Recovered $Q_1$"
    ), align = "lcccc",
    booktab = TRUE,
    position = "!h",
    escape = FALSE,
    caption = "\\label{tab:pi-model-sel}Model selection for varying $\\pi$ mixture parameters",
    format = "latex"
) %>% kableExtra::add_header_above(c(" "=1,"Models"=4,"Blocks"=1))
kable(
    (model_comparison_eps_rho %>%
        select(-one_of("n")) %>%
        pivot_wider(
            names_from = preferred_model,
            values_from = prop_model,
            values_fill = 0
        )  %>% group_by(epsilon_rho) %>% 
        summarise(rec_Q2 = meanse(rec_Q2), 
        iid = sum(iid), 
        pi = sum(pi), 
        rho = sum(rho), 
        pirho = sum(pirho)))[,c(1,3:6, 2)],
    digits = 2,
    col.names = c(
        "$\\eps[\\rho]$",
        "$iid\\text{-}colBiSBM$    ",
        "$\\pi\\text{-}colBiSBM$",
        "$\\rho\\text{-}colBiSBM$",
        "$\\pi\\rho\\text{-}colBiSBM$",
        "Recovered $Q_2$"
    ), align = "lcccc",
    booktab = TRUE,
    position = "!h",
    escape = FALSE,
    caption = "\\label{tab:rho-model-sel}Model selection for varying $\\rho$ mixture parameters",
    format = "latex"
) %>% kableExtra::add_header_above(c(" "=1,"Models"=4,"Blocks"=1))
```

\begin{figure}[H]
\includegraphics{./Rcodes/simulation/img/plot_model_function_eps.png}
\caption{Plot of preferred model in function of $\eps[\pi]$ and $\eps[\rho]$}
\label{fig:pref_model_func_eps}
\end{figure}

\paragraph{Results:}On the figure \ref{fig:pref_model_func_eps} and tables \ref{tab:pi-model-sel} 
and \ref{tab:rho-model-sel}, one can see that there is a turning
point around $\eps[\pi] = 0.2$ (resp. $\eps[\rho] = 0.2$), before which 
$iid\text{-}colBiSBM$
and $\rho\text{-}colBiSBM$ (resp. $\pi\text{-}colBiSBM$) are selected most of 
the times and after $0.2$ the $\pi\text{-}colBiSBM$ (resp.
$\rho\text{-}colBiSBM$) and 
$\pi\rho\text{-}colBiSBM$ gets more and more selected, highlighting our
capacity to recover the simulated structure.

\paragraph*{Remark:} Please note that when "Recovered $Q_1$(or $Q_2$)" is not an integer it's because
some procedures returned a value other than 3.